Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-03-03 manuscript/climate_sensitivity_draft.Rmd
reduced use of acronymns in text

Debates have also arisen regarding the traits influencing tree growth responses to drought. Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011](**Martin-Benito & Pederson 2015, doi:10.1111/jbi.12462**) and other forest biomes around the world [@greenwood_tree_2017]. However, in other cases these traits could not explain drought tolerance drought tolerance ([Maréchaux et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), or the direction of response was not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, their perceived influence on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011]. Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019]. These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

@@

Height measurements ($H$, n=1,518 trees) were taken by several researchers for a variety of purposes between 2012 to 2019. Measurement methods included manual [@stovall_assessing_2018; @neon_national_2018], digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018], and automatic, ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Measurements from the National Ecological Observatory Network (NEON) were collected near the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018]. Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H]$). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed by combining the height measurements across all species.

@@

Hydraulic traits were collected in August 2018 (Table 3). We sampled small sun-exposed branches up to eight meters above ground from three individuals of each species in and around the ForestGEO plot. Sampled branches were re-cut under water at least two nodes above the original cut and re-hydrated overnight in covered buckets (opaque plastic bags) before measurements were taken. Rehydrated leaves taken towards the apical end of the branch (n=3 per individual: small, medium, and large) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed. Leaf area was calculated from scanned images using the LeafArea R package [@R-LeafArea]. $LMA$ was calculated as the ratio of leaf dry mass to fresh area. $PLA_{dry}$ was calculated as the percent loss of area between fresh and dry leaves. $WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume, which was estimated using Archimedes’ displacement. We used the rapid determination method of @bartlett_rapid_2012 to estimate water potential at turgor loss point ($\pi_{tlp}$). Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA). Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012].

@@

Crown position varied as expected with $H$ (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. 2d). Crown position was a much poorer predictor of $Rt$ than was $H$ in the single-variable tests (Table 4), lending little overall support to the hypothesis that crown exposure reduces $Rt$ (Table 1). When considered alone, crown position had a significant response only in the 1966 drought, during which trees with dominant crown positiono had the lowest $Rt$. When $H$ was included in the model, crown position was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees. Crown position was included in some of the full models (Table 5). In 1977, where $H$ was not included in the full model, dominant trees had the lowest $Rt$, and suppressed the highest. In contrast, in full models including both height and crown position (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.

@@

Hydraulic traits, including xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$, were linked to drought responses (Tables 1,4,5). In the single-variable tests, $LMA$ and $WD$ were never significantly associated with $Rt$ (Table 4) and were excluded from the full models. In contrast, xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) during at least one of the three droughts (Table 4). $PLA_{dry}$ was a strong predictor for 1966 and all droughts combined, with consistently negative coefficients (Table 4). Similarly, $PLA_{dry}$ was consistently included, with negative coefficient, in full models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5). $\pi_{tlp}$ was not significant in any single-variable tests; however, coefficients were consistently negative (Table 4) and $\pi_{tlp}$ was included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5). Xylem porosity was not significant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5).

@@

Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height rather than crown exposure [@liu_effect_1993; @stovall_tree_2019]. We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ (consistent with @zuleta_drought-induced_2017; @stovall_tree_2019), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. The negative effect of height on $Rt$ held after accounting for species' traits.  Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem type (ring- vs. diffuse porous), but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality (**Sapes et al. 2019, doi: 10.1093/treephys/tpz062**).

>>>>

Debates have also arisen regarding the traits influencing tree growth responses to drought. Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011](**Martin-Benito & Pederson 2015, doi:10.1111/jbi.12462**) and other forest biomes around the world [@greenwood_tree_2017]. However, in other cases these traits could not explain drought tolerance drought tolerance ([Maréchaux et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), or the direction of response was not always consistent; for instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, their perceived influence on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011]. Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured wood density and $LMA$ [@medeiros_extensive_2019]. These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

@@

Tree heights ($H$) were measured by several researchers for a variety of purposes between 2012 to 2019 (n=1,518 trees). Measurement methods included manual [@stovall_assessing_2018; @neon_national_2018], digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018], and automatic, ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Measurements from the National Ecological Observatory Network (NEON) were collected near the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018]. Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H]$). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed by combining the height measurements across all species.

@@

Hydraulic traits were collected in August 2018 (Table 3). We sampled small sun-exposed branches up to eight meters above ground from three individuals of each species in and around the ForestGEO plot. Sampled branches were re-cut under water at least two nodes above the original cut and re-hydrated overnight in covered buckets (opaque plastic bags) before measurements were taken. Rehydrated leaves taken towards the apical end of the branch (n=3 per individual: small, medium, and large) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed. Leaf area was calculated from scanned images using the LeafArea R package [@R-LeafArea]. $LMA$ was calculated as the ratio of leaf dry mass to fresh area. $PLA_{dry}$ was calculated as the percent loss of area between fresh and dry leaves. wood density was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume, which was estimated using Archimedes’ displacement. We used the rapid determination method of @bartlett_rapid_2012 to estimate water potential at turgor loss point ($\pi_{tlp}$). Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA). Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012].

@@

Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. 2d). Crown position was a much poorer predictor of $Rt$ than was height in the single-variable tests (Table 4), lending little overall support to the hypothesis that crown exposure reduces $Rt$ (Table 1). When considered alone, crown position had a significant response only in the 1966 drought, during which trees with dominant crown positiono had the lowest $Rt$. When height was included in the model, crown position was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees. Crown position was included in some of the full models (Table 5). In 1977, where height was not included in the full model, dominant trees had the lowest $Rt$, and suppressed the highest. In contrast, in full models including both height and crown position (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.

@@

Hydraulic traits, including xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$, were linked to drought responses (Tables 1,4,5). In the single-variable tests, $LMA$ and wood density were never significantly associated with $Rt$ (Table 4) and were excluded from the full models. In contrast, xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) during at least one of the three droughts (Table 4). $PLA_{dry}$ was a strong predictor for 1966 and all droughts combined, with consistently negative coefficients (Table 4). Similarly, $PLA_{dry}$ was consistently included, with negative coefficient, in full models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5). $\pi_{tlp}$ was not significant in any single-variable tests; however, coefficients were consistently negative (Table 4) and $\pi_{tlp}$ was included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5). Xylem porosity was not significant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5).

@@

Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height rather than crown exposure [@liu_effect_1993; @stovall_tree_2019]. We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ (consistent with @zuleta_drought-induced_2017; @stovall_tree_2019), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. The negative effect of height on $Rt$ held after accounting for species' traits.  Drought sensitivity was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality (**Sapes et al. 2019, doi: 10.1093/treephys/tpz062**).

### 2020-03-03 manuscript/tables&figures.Rmd
reduced use of acronymns in text

```{r, Table 3, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table3 <- read.csv("tables_figures/table3_species_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
table3[8,8] <- paste0(table3[8,8], footnote_marker_symbol(1))
kable(table3, format = "latex", booktabs=TRUE, caption = "Table 3. Overview of analyzed species, listed in order of their relative contributions to woody stem productivity ($ANPP_{stem}$) in the plot, along with numbers and sizes sampled, and species traits.", escape=FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," ", "DBH (cm)" = 2, "species traits (mean +/- sd)" = 5)) %>%
  #kable_styling(bootstrap_options = "striped", full_width = FALSE)
kableExtra::footnote(symbol = c("Semi-ring porosity is intermediate between ring and diffuse. We group it with diffuse-porous species for more even division of species between categories."), footnote_as_chunk = FALSE)
```

\clearpage
```{r table 4, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
singlevar_table <- read.csv("tables_figures/table4_singlevariable_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
kable(singlevar_table, booktabs = TRUE, caption = "Table 4. Single-variable tests of hypothesized drivers of drought resistance. Models including each variable were compared to corresponding null models. dAIC is the AICc of the null model minus that of the model including the variable (thus, dAICc>2 indicates that the variable significantly improves the model", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
  pack_rows("Tree size and microenvironment", 1, 12, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
  pack_rows("Species traits", 13, 18, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
  kable_styling()
```
*dAICc > 1: variable qualified for inclusion in full model

>>>>

```{r, Table 3, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table3 <- read.csv("tables_figures/table3_species_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
table3[8,8] <- paste0(table3[8,8], footnote_marker_symbol(1))
kable(table3, format = "latex", booktabs=TRUE, caption = "Table 3. Overview of analyzed species, listed in order of their relative contributions to woody stem productivity ($ANPP_{stem}$) in the plot, along with numbers and sizes sampled, and species traits. Variable abbreviations are as in Table 2.", escape=FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," ", "DBH (cm)" = 2, "species traits (mean +/- sd)" = 5)) %>%
  #kable_styling(bootstrap_options = "striped", full_width = FALSE)
kableExtra::footnote(symbol = c("Semi-ring porosity is intermediate between ring and diffuse. We group it with diffuse-porous species for more even division of species between categories."), footnote_as_chunk = FALSE)
```

\clearpage
```{r table 4, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
singlevar_table <- read.csv("tables_figures/table4_singlevariable_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
kable(singlevar_table, booktabs = TRUE, caption = "Table 4. Single-variable tests of hypothesized drivers of drought resistance. Models including each variable were compared to corresponding null models. dAIC is the AICc of the null model minus that of the model including the variable (thus, dAICc>2 indicates that the variable significantly improves the model). Variable abbreviations are as in Table 2.", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
  pack_rows("Tree size and microenvironment", 1, 12, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
  pack_rows("Species traits", 13, 18, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
  kable_styling()
```
*dAICc > 1: variable qualified for inclusion in full model
